Code Repository

### 2025-06-15 Code Repository.Rmd
Updated Latent variable code
Latent variable code added

@@

## Pre-processing
```{r}
library(dplyr)
library(MASS)
library(tidyr)
library(mvtnorm)
library(ggplot2)
library(kableExtra)

@@

Creating permuted null distributions using permutation function
```{r}
set.seed(24253739)
permuted_null_distributions <-permutation_function(tmpDat$gw32, tmpDat$gw38, tmpDat$binary_insulin)
permuted_null_distributions.df <-data.frame(permuted_null_distributions)
write.csv(permuted_null_distributions.df, "Permuted Null Distributions.csv", row.names = FALSE)

@@

## Latent Variable Analysis McMenamin et al.
```{r}

@@

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = FALSE,
  message = FALSE,
  warning = FALSE)

@@

## Pre-processing
```{r}
library(dplyr)
library(MASS)
library(tidyr)
library(mvtnorm)
library(ggplot2)
library(kableExtra)
library(stats)
library(nlme)
library(boot)
library(matrixcalc)
library(numDeriv)
library(cubature)
library(optimx)
library(brglm)
library(Matrix)
library(matrixcalc)

@@

Creating permuted null distributions using permutation function
```{r}
set.seed(24253739)
permuted_null_distributions <-permutation_function(tmpDat$gw32, tmpDat$gw38, tmpDat$binary_insulin)
permuted_null_distributions.df <-data.frame(permuted_null_distributions)
write.csv(permuted_null_distributions.df, "Permuted Null Distributions.csv", row.names = FALSE)
p_value_squared_composite <- sum(permuted_null_distributions.df$permutation_squared_composite >= study_squared_composite)/100000
p_value_summed_composite <- sum(permuted_null_distributions.df$permutation_summed_composite <= study_summed_composite)/100000
print(paste("P-value squared composite: ", p_value_squared_composite))
print(paste("P-value summed composite: ", p_value_summed_composite))
```

@@

## Latent Variable Analysis McMenamin et al.
The following code is from McMenamin et al. 2021 which has been modified to remove the code for baseline variables of the continuous outcomes.
```{r}
#Likelihood function

f<-function(X,dat)
{
  #data
  #dat<-dat[!is.na(dat[,3]),]#change this to baseline measure when it is added?

  #parameters
  alpha0 <- X[1]
    alpha1 <- X[2]
       beta0 <- X[3]
         beta1 <- X[4]
           psi0 <- X[5]
             psi1 <- X[6]

  #covariance parameters
  sig1 <- exp(X[7]) 
    sig2 <- exp(X[8]) 
     rho12 <- 2*inv.logit(X[9])-1
       rho13 <- 2*inv.logit(X[10])-1
          rho23 <- 2*inv.logit(X[11])-1

  #Known cutoffs
  tau0 <- -Inf
   tau1 <- 0
     tau2 <- +Inf

  print(X)

  #model means
  muz1<-alpha0+alpha1*dat[,2]
   muz2<-beta0+beta1*dat[,2]
     muz3<-psi0+psi1*dat[,2]

  #conditional means
  muz3cond <- muz3+((rho13-rho12*rho23)*(dat[,3]-muz1)/(sqrt(sig1)*(1-(rho12)^2)))+((rho23-rho13*rho12)*(dat[,4]-muz2)/(sqrt(sig2)*(1-(rho12)^2)))

  #conditional covariance
  sigcond <- 1-(((rho13)^2-2*rho12*rho13*rho23+(rho23)^2)/(1-(rho12)^2))

  #continuous bivariate covariance
  matbiv11 <- (sig1)^2
   matbiv12 <- rho12*sig1*sig2
    matbiv22 <- (sig2)^2
      Sigbiv   <- matrix(c(matbiv11, matbiv12, matbiv12, matbiv22), nrow=2, ncol=2)#check
        Sigbiv <- (Sigbiv*t(Sigbiv))^0.5

  #upperlimits
  mulim0<-matrix( tau1-muz3cond, ncol=1)
   mulim1<-matrix( tau2-muz3cond, ncol=1)

  #binprob
  pr30<-apply(mulim0,1,function(x){return(pmvnorm(lower=-Inf,upper=x,mean=0,sigma = sigcond))})
   pr31<-apply(mulim1,1,function(x){return(pmvnorm(lower=-Inf,upper=x,mean=0,sigma = sigcond))})
     prz12<-dmvnorm(cbind(dat[,3],dat[,4]), c(mean(muz1), mean(muz2)), Sigbiv)

  #Likelihood

  #components of likelihood,k=0,1 (binary)
  l0<-log(pr30)+log(prz12)#k=0
  l1<-log(pr31-pr30)+log(prz12)#k=1

  data0 <- cbind(dat[,5],l0)#0
   data1 <- cbind(dat[,5],l1)#1

  #0
  data0[data0[,1]==1,2]<-0

  #1
  data1[data1[,1]==0,2]<-0

  t0 <- sum(data0[,2])
    t1 <- sum(data1[,2])

  #-log(likelihood)
  Tfinal<-sum(t0)+sum(t1)

  return(-Tfinal)
}

lowerlim <- c(-Inf,-Inf,-Inf,-Inf,-Inf,-Inf,-Inf,-Inf,-Inf,-Inf,-Inf)
upperlim <- c(+Inf,+Inf,+Inf,+Inf,+Inf,+Inf,+Inf,+Inf,+Inf,+Inf,+Inf)

##Probability of response
integrand<-function(Zint,meantreat,meanuntreat,mle)
{

  sigmahat=matrix(nrow=3,ncol=3)
    sigmahat[1,1]=(exp(mle[7]))^2
     sigmahat[2,1]=(2*inv.logit(mle[9])-1)*(exp(mle[7]))*exp(mle[8])
       sigmahat[3,1]=(2*inv.logit(mle[10])-1)*(exp(mle[7]))
           sigmahat[1,2]=sigmahat[2,1]
             sigmahat[2,2]=(exp(mle[8]))^2
               sigmahat[3,2]=(2*inv.logit(mle[11])-1)*(exp(mle[8]))
                 sigmahat[1,3]=sigmahat[3,1]
                   sigmahat[2,3]=sigmahat[3,2]
                     sigmahat[3,3]=1

  xtreat<-cbind(-meantreat[,1]+Zint[1], -meantreat[,2]+Zint[2], -meantreat[,3]+Zint[3])
    xuntreat<-cbind(-meanuntreat[,1]+Zint[1],-meanuntreat[,2]+Zint[2],-meanuntreat[,3]+Zint[3])

  pdftreat=dmvnorm(xtreat, mean=c(0,0,0),sigma=sigmahat)
     pdfuntreat=dmvnorm(xuntreat, mean=c(0,0,0),sigma=sigmahat)

  return(c(mean(pdftreat),mean(pdfuntreat)))
}

probofsuccess<-function(mle,n,dat,eta)
{
  n=n

  meantreat=cbind(cbind(rep(1,n),rep(1,n))%*%c(mle[1:2]),cbind(rep(1,n),rep(1,n))%*%c(mle[3:4]), 
                  cbind(rep(1,n),rep(1,n))%*%mle[5:6])   
  meanuntreat=cbind(cbind(rep(1,n),rep(0,n))%*%c(mle[1:2]),cbind(rep(1,n),rep(0,n))%*%c(mle[3:4]), 
                    cbind(rep(1,n),rep(0,n))%*%mle[5:6])

  #lower and upper bounds
  minmean1=min(c(meantreat[,1],meanuntreat[,1]))
  minmean2=min(c(meantreat[,2],meanuntreat[,2]))
  minmean3=min(c(meantreat[,3],meanuntreat[,3]))

  maxmean1=max(c(meantreat[,1],meanuntreat[,1]))
  maxmean2=max(c(meantreat[,2],meanuntreat[,2]))
  maxmean3=max(c(meantreat[,3],meanuntreat[,3]))

  lower=c(qnorm(1e-15,minmean1,exp(mle[7])),qnorm(1e-15,minmean2,exp(mle[8])),qnorm(1e-15,minmean3,1))

  upper=c(eta[1],eta[2],0)

  a=cuhre(f=integrand,nComp=2,lower=lower,upper=upper,flags=list(verbose=0,final=1,pseudo.random=0,mersenne.seed=NULL),
          meantreat=meantreat,meanuntreat=meanuntreat,mle=mle)
  #return(c(a$value[1],a$value[2]))
  #return(c(a$value[1]-a$value[2],a$value[1],a$value[2]))
  return(c((log(a$integral[1]/(1-a$integral[1]))-log(a$integral[2]/(1-a$integral[2]))),log(a$integral[1]/a$integral[2]),
           a$integral[1]-a$integral[2],a$integral[1],a$integral[2]))
  #return(log(a$value[1]/a$value[2]))
}

#Partial derivatives

partials<-function(mle,n,dat,eta)
{
  p=length(mle)
  fit1<-probofsuccess(mle,n,dat,eta)
  fitOR<-fit1[1]
  fitRR<-fit1[2]
  fitRD<-fit1[3]
  partials.augbinOR<-as.vector(rep(0,p))
  partials.augbinRR<-as.vector(rep(0,p))
  partials.augbinRD<-as.vector(rep(0,p))

  for(i in 1:p){
    valueupdate=mle
    valueupdate[i]=valueupdate[i]+0.000001

    updateprobOR=probofsuccess(valueupdate,n,dat,eta)[1]
    updateprobRR=probofsuccess(valueupdate,n,dat,eta)[2]
    updateprobRD=probofsuccess(valueupdate,n,dat,eta)[3]

    partials.augbinOR[i]=(updateprobOR-fitOR)/0.000001
    partials.augbinRR[i]=(updateprobRR-fitRR)/0.000001
    partials.augbinRD[i]=(updateprobRD-fitRD)/0.000001

  return(c(partials.augbinOR,partials.augbinRR,partials.augbinRD,fit1))
}

result.latent<-as.list(NULL)
results<-as.list(NULL)

#Function for treatment effects and CIs for both methods

LatVarfunc<-function(dat,eta){
  n=dim(dat)[1]

  #Starting values
  lm1<-lm(dat[,3]~dat[,2])
  lm2<-lm(dat[,4]~dat[,2])
  lm3<-lm(dat[,5]~dat[,2])
  sig1est<-log(sqrt(var(dat[,3])))
  sig2est<-log(sqrt(var(dat[,4])))
  rho12est<-log(((cor(dat[,3],dat[,4])+1)/2)/(1-(cor(dat[,3],dat[,4])+1)/2))
  rho13est<-log(((cor(dat[,3],dat[,5])+1)/2)/(1-(cor(dat[,3],dat[,5])+1)/2))
  rho23est<-log(((cor(dat[,4],dat[,5])+1)/2)/(1-(cor(dat[,4],dat[,5])+1)/2))

  X<-c(lm1$coef[1],lm1$coef[2],lm2$coef[1],lm2$coef[2],lm3$coef[1],lm3$coef[2],sig1est,sig2est,rho12est, rho13est,rho23est)
  X<-as.vector(X)

  #Latent variable

  mlefit=optimx(X,f,dat=dat,lower=lowerlim,upper=upperlim,method="nlminb",control=list(rel.tol=1e-12))
  mle<-coef(mlefit[1,])
  hess<-attr(mlefit,"details")["nlminb",]$nhatend
  print(hess)
  mlecov=ginv(hess)
  mlecov<-nearPD(mlecov)$mat
  se<-sqrt(diag(mlecov))
  print(se)

  part<-partials(mle,n,dat,eta)
  print(part)
  meanOR<-part[34]

  partsOR <- part[1:11]
  varianceOR=t(partsOR)%*%mlecov%*%partsOR
  varianceOR=varianceOR[1,1]

  meanRR<-part[35]
  partsRR<-part[12:22]
  varianceRR=t(partsRR)%*%mlecov%*%partsRR
  varianceRR=varianceRR[1,1]

  meanRD<-part[36]
  partsRD<-part[23:33]
  varianceRD=t(partsRD)%*%mlecov%*%partsRD
  varianceRD=varianceRD[1,1]

  CIOR<-c(meanOR-1.96*sqrt(varianceOR),meanOR,meanOR+1.96*sqrt(varianceOR))
  CIRR<-c(meanRR-1.96*sqrt(varianceRR),meanRR,meanRR+1.96*sqrt(varianceRR))
  CIRD<-c(meanRD-1.96*sqrt(varianceRD),meanRD,meanRD+1.96*sqrt(varianceRD))

  probresplat<-c(part[37],part[38])
  result.latent<-c(CIOR,CIRR,CIRD,probresplat)

  epshat1 <- dat[,3]-(mle[1]+mle[2]*dat[,2])
  epshat2 <- dat[,4]-(mle[3]+mle[4]*dat[,2])
  epshat3 <- dat[,5]-(mle[5]+mle[6]*dat[,2])
  epshat <- as.matrix(cbind(epshat1,epshat2,epshat3))

  sig1hat <- exp(mle[7]) 
  sig2hat <- exp(mle[8]) 
  ##Correlation coefficients (rho12, rho13, rho23) must lie in the interval (-1, 1). However, optimization routines like optimx work best on    unconstrained parameters, ranging from -Inf to +Inf. To enable this, the code uses the logistic (sigmoid) transformation to map an           unconstrained value to the range (0,1), and then linearly transforms it to (−1,1).##
  rho12hat <- 2*inv.logit(mle[9])-1
  rho13hat <- 2*inv.logit(mle[10])-1
  rho23hat <- 2*inv.logit(mle[11])-1

  SigHat<-matrix(data=c(sig1hat^2,rho12hat*sig1hat*sig2hat,rho13hat*sig1hat,rho12hat*sig1hat*sig2hat,sig2hat^2,
           rho23hat*sig2hat,rho13hat*sig1hat,rho23hat*sig2hat,1),ncol=3)
  modres <- diag((epshat)%*%ginv(SigHat)%*%t(epshat))

  results<-c(result.latent,modres,mle,se)
  results1 <- c(result.latent, mle, se)
  return(results1)
}
```

The output of the latvarfunc is as follows:
1	logOR_LCI	Lower 95% CI for the log odds ratio
2	logOR	Point estimate for the log odds ratio
3	logOR_UCI	Upper 95% CI for the log odds ratio
4	logRR_LCI	Lower 95% CI for the log relative risk
5	logRR	Point estimate for the log relative risk
6	logRR_UCI	Upper 95% CI for the log relative risk
7	RD_LCI	Lower 95% CI for the risk difference
8	RD	Point estimate for the risk difference
9	RD_UCI	Upper 95% CI for the risk difference

Reprocessing tmpDat into the suitable format "inverted_data" for the LatVarfunc
Requires setting columns in a specific order and inverting the data
Column 1 is patient ID
Column 2 is treatment (binary 0,1)
Column 3 is gw32
Column 4 is gw38
Column 5 is Insulin (binary 0, 1)
Make sure to remove ## for the code to run
```{r}
data <- tmpDat[,c("Label", "binary_group", "gw32", "gw38", "binary_insulin")]
inverted_data <- data[complete.cases(data), ]
inverted_data$binary_insulin <- ifelse(inverted_data$binary_insulin == 1, 0, 1)
inverted_data$gw32 <- inverted_data$gw32 * -1
inverted_data$gw38 <- inverted_data$gw38 * -1
inverted_eta <- c(-5.1, -5.1) ## these are the "cut-off" points for the continuous variables
results <- LatVarfunc(inverted_data, inverted_eta)
names(results) <- c("logOR_LCI", "logOR", "logOR_UCI")
```

Power estimation of Latent variable analysis
*Note that this takes a long time to run, and as such we only ran it for two sample sizes, 100 and 150*
```{r}
set.seed(24253739)
power_function_null_dist <- function(n) {
  power_result <- matrix(NA, nrow = 100, ncol = 3) 
  i <- 1
  repeat {
    placebo_group <- inverted_data[inverted_data$binary_group == 0, ]
    sampled_placebo <- placebo_group[sample(nrow(placebo_group), n/2, replace = TRUE), ]

    metformin_group <- inverted_data[inverted_data$binary_group == 1, ]
    sampled_metformin <- metformin_group[sample(nrow(metformin_group), n/2, replace = TRUE), ]

    sampled_data <- rbind(sampled_placebo, sampled_metformin) ##Needed for latvarfunc?
    inverted_eta <- c(-5.1, -5.1)

    LatVarfunc_result1<- LatVarfunc(sampled_data, inverted_eta)

    power_result[i, ] <- LatVarfunc_result1[1:3]
    i<- i +1
    if(i > 100) break}
  return(sum(power_result[,1] > 0 | power_result[,3] < 0)/ 100)
}
sample_vector <- 100
##null_dist_summed_TS_results <- sapply(sample_vector, function(n_value) {power_function_null_dist(n_value)})
power_100_latvar<-0.99
power_150_latvar <- 0.99
```

Type I error rate estimation of LatVarfunc
*Takes a long time to run this, only ran it for sample sizes of 100 and 200*
```{r}
set.seed(24253739)
type_1_error_func <- function(n) {
  power_result <- matrix(NA, nrow = 100, ncol = 3) 
  i <- 1
  repeat {

    sampled_data <- inverted_data[sample(nrow(inverted_data), n, replace = TRUE), ]
    sampled_data$binary_group <- sample(rep(c(0, 1), each = n / 2))

    inverted_eta <- c(-5.1, -5.1)

    LatVarfunc_result1<- LatVarfunc(sampled_data, inverted_eta)

    power_result[i, ] <- LatVarfunc_result1[1:3]
    i<- i +1
    if(i > 100) break}
  return(sum(power_result[,1] > 0 | power_result[,3] < 0)/ 100)
}
sample_vector <- 200
type_1_error <- sapply(sample_vector, function(n_value) { type_1_error_func(n_value)})

type_1_error_rate_100 <- 0.49
type_1_error_rate_200 <- 0.57
```

P-value Estimation of LatVarFunc
```{r}
rd <- -0.043
LCI <- -0.0488
UCI <- -0.0372

se <- (rd-LCI)/1.96
to_get_zero <- rd/se
p_value_latvar <-2*(1-pnorm(abs(to_get_zero)))
```

@@

## References
 McMenamin, Martina, Jessica K Barrett, Anna Berglind, and James Ms Wason. 2021. “Employing a Latent
 Variable Framework to Improve Efficiency in Composite Endpoint Analysis.” Statistical Methods in
 Medical Research 30 (3): 702–16. https://doi.org/10.1177/0962280220970986.
